---
title: "Prediction Assignment"
author: "Lucas Antelo"
date: "31 Mai 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Executive Summary


## Data retrieval and processing

```{r cars, cache=TRUE, results='hide'}
urlTrainingData <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
urlTestingData <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
date_downloaded <- date()
download.file(urlTrainingData, destfile = "training.csv")
download.file(urlTestingData, destfile = "testing.csv")
training <- read.csv("training.csv",stringsAsFactors = FALSE)
testing <- read.csv("testing.csv", stringsAsFactors = FALSE)
rm("urlTrainingData", "urlTestingData") #remove variables to free memory space
```

## Exploratory Data Analysis

You can also embed plots, for example:

```{r pressure, echo=FALSE}
dim(training)
summary(training)

completness <- sapply(training, complete.cases) #T/F for each element
colCompleteness <- apply(completness,2,sum) #sum of the logical values 1/0 columnwise

length(which(colCompleteness < max(colCompleteness))) 
#67 columns are less than 19622 and thus dataset will be reduced to complete cases

trainingSubset <- training[complete.cases(training),]#dim 406 rows and 60 columns
trainingClean <- cbind(trainingSubset[,8:13], trainingSubset[,18:24], trainingSubset[,37:48], trainingSubset[,60:71], trainingSubset[, 81:88])
trainingClean <- cbind(trainingClean, trainingSubset[,99:100], trainingSubset[,113:126], trainingSubset[,137:139], trainingSubset[,151:160])# dim 406 to 74
#get rid of the unnecessary columns
trainNumeric <- sapply(training[,2:dim(training)[2]], as.numeric) #especially strings are transformed into NAs
training <- cbind(training[,1], trainNumeric)

```
The training dataset contains variables with a devastating number of non available numbers or incompleteness. The number of complete cases diminishes to 406, which represents 2% of the raw dataset. The question remains, if I use all variables for my prediction model, especially those, with many NA values.

